---
title: "Project - Group 4"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

[Link to the instructions of the project](https://owncloud.ulb.ac.be/index.php/s/an85YFknNiDQXGA?path=%2Fresources#pdfviewer)  
[Link to the information of the variables name](https://ftp.ncbi.nlm.nih.gov/dbgap/studies/phs000424/phs000424.v8.p2/pheno_variable_summaries/phs000424.v8.pht002742.v8.GTEx_Subject_Phenotypes.data_dict.xml)   
Technical variable:  
- COHORT(categorical)    
- TRISCHD(numerical)   
- DTHHRDY(categorical)    
- DTHVNT(categorical)    
Demographic/Health:  
- AGE(numerical)   
- SEX(categorical)       
- HGHT(numerical)   
- WGHT(numerical)   
- BMI(numerical)   


#Question 3

```{r}
#Installation of the Bioconductor + Installation of the DESeq2
if (!requireNamespace('BiocManager', quietly = TRUE))
    install.packages('BiocManager')
BiocManager::install("DESeq2")
```


```{r}
#Import all libraries
library(readr)
library(ggplot2)
library(cowplot)
library(DESeq2)
library(dplyr) #install.packages("dplyr")
library(pheatmap)
library(tidyr)
library(tibble)
library(reshape2)
```


```{r}
# Path to file (raw data)
clinical_path <- "./TSV_files/raw/clinical_data.tsv" #Your path here
df_clinical<- read.delim(clinical_path)

morphological_path <- "./TSV_files/raw/morphological_counts_lunit_dino.tsv" #Your path here
df_morphological<- read.delim(morphological_path)

rna_path <- "./TSV_files/raw/RNA_read_counts.tsv" #Your path here
df_rna<- read.delim(rna_path)


# Path to file (Already processed data obtained in the previous questions)
rna_filtered_path <- "./TSV_files/General/df_rna_filtered.tsv" #Your path here
df_rna_filtered <- read.table(rna_filtered_path, header = TRUE, sep = "\t")

rna_transposed_path <- "./TSV_files/General//df_rna_transposed.tsv" #Your path here
transposed_df <- read.table(rna_transposed_path, header = TRUE, sep = "\t")
```



```{r}
print(colnames(transposed_df))
```


Cleaning process
```{r}
#Cleaning PROCRESS:

#Start with 56200 genes in total
#First filter, genes that has a sum less than 10 is removed from the dataset (45675 genes left)

df_rna_clean <- df_rna %>%
  mutate(across(3:ncol(df_rna), as.numeric)) %>%
  mutate(sum_expression = rowSums(across(3:ncol(df_rna)), na.rm = TRUE)) %>%
  filter(sum_expression > 10) %>%
  select(-sum_expression)

str(df_rna_clean)



#Second filter: calcul the MAD value for each gene (Median Absolute Deviation) because we want to keep the genes with the highest deviation and variability. In this case, we are going to use the log10(MAD +1), as explained in the project question as hint. Then, based on the histogram, a threshold is found (Q3) to separate the histogram into 2 different groups, with the highest and lowest value, to remove all the genes with the lowest variations. At the end, we only have 11382 genes left.

median_expression <- apply(df_rna_clean[, 3:ncol(df_rna_clean)], 1, median, na.rm = TRUE)
mad_values <- apply(df_rna_clean[, 3:ncol(df_rna_clean)], 1, function(x) {
  median(abs(x - median(x, na.rm = TRUE)), na.rm = TRUE)
})

df_rna_clean$mad_value <- mad_values

df_rna_clean$log10_mad_plus_1 <- log10(df_rna_clean$mad_value + 1)

threshold_mad_log10 <- quantile(df_rna_clean$log10_mad_plus_1, 0.75, na.rm = TRUE)

print(dim(df_rna_clean))
print(head(df_rna_clean))

histogram_plot <- ggplot(df_rna_clean, aes(x = log10_mad_plus_1)) +
  geom_histogram(binwidth = 0.5, fill = "skyblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of log10(MAD + 1)",
       x = "log10(MAD + 1)",
       y = "Frequency") +
  geom_vline(aes(xintercept = threshold_mad_log10), color = "red", linetype = "dashed") +
  geom_text(aes(x = threshold_mad_log10, y = Inf, label = paste("Threshold =", round(threshold_mad_log10, 2))), 
            color = "red", angle = 90, vjust = 1.5, hjust = -0.5)

print(histogram_plot)

df_rna_high_variability <- df_rna_clean %>%
  filter(log10_mad_plus_1 > threshold_mad_log10) %>%
  select(-log10_mad_plus_1)

print(dim(df_rna_high_variability))
print(head(df_rna_high_variability))

# write.table(df_rna_high_variability, file = "df_rna_filtered.tsv", sep = "\t", row.names = FALSE) #Uncomment to save it

```

Transposition et manipulation dataframe pour lancer l'analyse DeSeq2
```{r}
transposed_df <- as.data.frame(t(df_rna_filtered[, 1:ncol(df_rna_filtered)]))
transposed_df <- cbind(SMPLID = colnames(df_rna_filtered[, 1:ncol(df_rna_filtered)]), transposed_df)
colnames(transposed_df) <- transposed_df[2, ]

colnames(transposed_df)[1] <- "SMPLID"

transposed_df <- transposed_df[-c(1, 2), ]
transposed_df <- transposed_df[-nrow(transposed_df), ]
rownames(transposed_df) <- NULL
#write.table(transposed_df, file = "df_rna_transposed.tsv", sep = "\t", row.names = FALSE) #Uncomment to save it
```

Run l'analyse DeSeq sur chaque gene (11382) pour chaque cluster
```{r}
run_deseq <- function(formula, count_matrix, col_data) {
  print(formula)
  
  non_na_indices <- complete.cases(col_data)
  filtered_count_matrix <- count_matrix[, non_na_indices, drop = FALSE]
  filtered_col_data <- col_data[non_na_indices, , drop = FALSE]
  
  if (ncol(filtered_count_matrix) == 0 || nrow(filtered_col_data) == 0) {
    stop("Erreur : la matrice de comptage ou les métadonnées sont vides après filtrage des NA.")
  }
  
  dds <- DESeqDataSetFromMatrix(countData = filtered_count_matrix, colData = filtered_col_data, design = as.formula(formula))
  dds <- DESeq(dds)
  
  print(results(dds))
  return(results(dds))
}

clinical_data <- transposed_df
morphological_data <- read_tsv(morphological_path, show_col_types = FALSE)

count_matrix <- t(as.matrix(morphological_data[, -1]))
colnames(count_matrix) <- morphological_data$SMPLID
rownames(count_matrix) <- colnames(morphological_data)[-1]

variables <- colnames(clinical_data)[-1]
results_list <- list()
skipped_genes <- list()

for (var in variables) {
  col_data <- DataFrame(col = clinical_data[[var]])
  names(col_data) <- var
  rownames(col_data) <- clinical_data$SMPLID
  
  if (nrow(col_data) == 0 || sum(complete.cases(col_data)) == 0) {
    skipped_genes[[length(skipped_genes) + 1]] <- var
  } else {
    res <- tryCatch({
      run_deseq(paste0("~ ", var), count_matrix, col_data)
    }, error = function(e) {
      message("Failed to run DESeq2 for variable: ", var, "; Error: ", e$message)
      NULL
    })

    if (!is.null(res) && !any(is.na(res$padj))) {
      results_list[[var]] <- res
    } else {
      skipped_genes[[length(skipped_genes) + 1]] <- var
    }
  }
}

dir.create("Cluster_2001_tot", showWarnings = FALSE)

for (cluster_idx in 0:31) {
  cluster_results <- data.frame(matrix(ncol = 7, nrow = 0))
  colnames(cluster_results) <- c("Genes", "BaseMean", "log2FoldChange", "lfcSE", "stat", "pvalue", "padj")
  
  for (gene in variables) {
    gene_results <- results_list[[gene]][cluster_idx + 1, ]
    
    row_values <- c(
      gene,
      if(length(gene_results$baseMean) > 0) gene_results$baseMean else NA,
      if(length(gene_results$log2FoldChange) > 0) gene_results$log2FoldChange else NA,
      if(length(gene_results$lfcSE) > 0) gene_results$lfcSE else NA,
      if(length(gene_results$stat) > 0) gene_results$stat else NA,
      if(length(gene_results$pvalue) > 0) gene_results$pvalue else NA,
      if(length(gene_results$padj) > 0) gene_results$padj else NA
    )
    
    temp_df <- as.data.frame(t(row_values), stringsAsFactors=FALSE)
    colnames(temp_df) <- c("Genes", "BaseMean", "log2FoldChange", "lfcSE", "stat", "pvalue", "padj")
    cluster_results <- rbind(cluster_results, temp_df)
  }
  
  #write.table(cluster_results, file = sprintf("Cluster_2001_tot/Cluster%02d_results.tsv", cluster_idx), sep = "\t", row.names = FALSE, col.names = TRUE, quote = FALSE) #Uncomment to save it
}



```

Permet d'assembler les fichiers créés en 1 seul fichier (l'analyse précédente a été découpé en plusieurs parties pour pas run tout en 1 fois)
```{r}
folders <- c("Cluster_1_1000", "Cluster_1001_2000", "Cluster_2001_tot")

for (i in 0:31) {
  
  cluster_file_name <- sprintf("Cluster%02d_results.tsv", i)
  
  temp_data_list <- lapply(folders, function(folder) {
    file_path <- file.path(folder, cluster_file_name)
    if (file.exists(file_path)) {
      
      read.table(file_path, header = TRUE, sep = "\t")
    } else {
      print("gfgggg")
      NULL  
    }
  })
  
  temp_data_list <- Filter(Negate(is.null), temp_data_list)
  
  if (length(temp_data_list) > 0) {
    combined_data <- do.call(rbind, temp_data_list)
    
    write.table(combined_data, sprintf("Cluster_tot/Cluster%02d_results.tsv", i), sep = "\t", row.names = FALSE, col.names = TRUE, quote = FALSE)
  }
}



```

Filtrage des valeurs NA obtenus lors de l'analyse (before: 11382 et after: 10279)
```{r}
input_folder <- "Cluster_tot"
output_folder <- "Cluster_filtered"

if (!dir.exists(output_folder)) {
  dir.create(output_folder)
}

files <- list.files(input_folder, pattern = "\\.tsv$", full.names = TRUE)

for (file_path in files) {
  data <- read.table(file_path, header = TRUE, sep = "\t")

  cat(sprintf("Taille avant filtrage de %s: %d lignes\n", basename(file_path), nrow(data)))

  data_clean <- na.omit(data)

  cat(sprintf("Taille après filtrage de %s: %d lignes\n", basename(file_path), nrow(data_clean)))

  output_file_path <- file.path(output_folder, basename(file_path))

  write.table(data_clean, output_file_path, sep = "\t", row.names = FALSE, col.names = TRUE, quote = FALSE)
}


```

Output matrix pour montrer le nombre de genes significatifs par cluster
```{r}
input_folder <- "Cluster_filtered" 

files <- list.files(input_folder, pattern = "\\.tsv$", full.names = TRUE)
files <- sort(files)

result_matrix <- matrix(nrow = length(files), ncol = 2, dimnames = list(NULL, c("Cluster", "Significant_Genes")))

for (i in seq_along(files)) {
  file_path <- files[i]
  data <- read.table(file_path, header = TRUE, sep = "\t")

  significant_genes_count <- sum(data$padj < 0.05, na.rm = TRUE) #Value below 0.05 for padj
  cluster_number <- as.numeric(sub(".*Cluster(\\d{2}).*", "\\1", basename(file_path)))
  result_matrix[i, ] <- c(cluster_number, significant_genes_count)
}

print(result_matrix)

```

Calcul moyenne des rna count (not used)
```{r}
genes_file_path <- "./TSV_files/Cluster_filtered/Cluster00_results.tsv"  #Your path here
genes_list <- read.table(genes_file_path, header = TRUE, sep = "\t")

gene_names <- genes_list$Genes  
results_df <- data.frame(Gene = character(), Mean = numeric(), stringsAsFactors = FALSE)

all_genes_from_df_rna <- unique(df_rna$Description)

gene_indices <- match(gene_names, all_genes_from_df_rna)

df_rna_filtered <- df_rna[gene_indices, ]

for (i in seq_along(gene_names)) {
  gene <- gene_names[i]
  gene_row <- df_rna_filtered[i, ]

  if (!is.na(gene_row$Description) && nrow(gene_row) > 0) {
    mean_value <- mean(as.numeric(gene_row[, 3:ncol(gene_row)]), na.rm = TRUE)
    results_df <- rbind(results_df, data.frame(Gene = gene, Mean = mean_value))
  } else {
    results_df <- rbind(results_df, data.frame(Gene = gene, Mean = NA))
  }
}

print(results_df)
write.table(results_df, "Mean_Gene.tsv", sep = "\t", row.names = FALSE, quote = FALSE)


```

Matrix UP/DOWN regulated genes
```{r}
log2fc_threshold <- 0
pval_threshold <- 0.05

input_folder <- "./TSV_files/Cluster_filtered" #Your path here

files <- list.files(input_folder, pattern = "\\.tsv$", full.names = TRUE)
files <- sort(files)

result_matrix <- matrix(nrow = length(files), ncol = 3, dimnames = list(NULL, c("Cluster", "Up_regulated_Genes", "Down_regulated_Genes")))

for (i in seq_along(files)) {
  file_path <- files[i]
  data <- read.table(file_path, header = TRUE, sep = "\t")

  up_regulated_count <- sum(data$log2FoldChange > log2fc_threshold & data$padj < pval_threshold, na.rm = TRUE)
  
  down_regulated_count <- sum(data$log2FoldChange < -log2fc_threshold & data$padj < pval_threshold, na.rm = TRUE)

  cluster_number <- as.numeric(sub(".*Cluster(\\d{2}).*", "\\1", basename(file_path)))

  result_matrix[i, ] <- c(cluster_number, up_regulated_count, down_regulated_count)
}

print(result_matrix)

```

Creation matrix top regulated genes
```{r}
log2fc_threshold <- 0
pval_threshold <- 0.05

input_folder <- "./TSV_files/Cluster_filtered" #Your path here
files <- list.files(input_folder, pattern = "\\.tsv$", full.names = TRUE)
files <- sort(files)

genes_file_path <- "./TSV_files/Cluster_filtered/Cluster00_results.tsv" #Your path here
genes_list <- read.table(genes_file_path, header = TRUE, sep = "\t")
unique_genes <- unique(genes_list$Genes)  # Ensure uniqueness

gene_df <- data.frame(Gene = unique_genes)

for (file_path in files) {
    data <- read.table(file_path, header = TRUE, sep = "\t")
    
    up_regulated <- data[data$log2FoldChange > log2fc_threshold & data$padj < pval_threshold, ]

    cluster_name <- sub(".*Cluster(\\d{2}).*", "\\1", basename(file_path))
    gene_df[[paste0("Cluster_", cluster_name)]] <- 0  

    if (length(up_regulated) > 0) {
        gene_df[[paste0("Cluster_", cluster_name)]] <- ifelse(gene_df$Gene %in% up_regulated$Genes, 1, 0)
    }
}

gene_df$Total_Up_Regulated <- rowSums(gene_df[, -1])

top_genes_df <- gene_df[order(-gene_df$Total_Up_Regulated), ]
#top_10_genes <- head(top_genes_df, 10)
#print(top_10_genes)

write.table(top_genes_df, "Top_Upregulated_Genes.tsv", sep = "\t", row.names = FALSE, quote = FALSE)

```

Creation matrix down regulated genes
```{r}
log2fc_threshold <- 0
pval_threshold <- 0.05

input_folder <- "./TSV_files/Cluster_filtered" #Your path here
files <- list.files(input_folder, pattern = "\\.tsv$", full.names = TRUE)
files <- sort(files)

genes_file_path <- "./TSV_files/Cluster_filtered/Cluster00_results.tsv" #Your path here
genes_list <- read.table(genes_file_path, header = TRUE, sep = "\t")
unique_genes <- unique(genes_list$Genes)  

gene_df <- data.frame(Gene = unique_genes)

for (file_path in files) {
    data <- read.table(file_path, header = TRUE, sep = "\t")
    
    down_regulated <- data[data$log2FoldChange < log2fc_threshold & data$padj < pval_threshold, ]

    cluster_name <- sub(".*Cluster(\\d{2}).*", "\\1", basename(file_path))
    gene_df[[paste0("Cluster_", cluster_name)]] <- 0  

    if (length(down_regulated) > 0) {
        gene_df[[paste0("Cluster_", cluster_name)]] <- ifelse(gene_df$Gene %in% down_regulated$Genes, 1, 0)
    }
}

gene_df$Total_Up_Regulated <- rowSums(gene_df[, -1])

top_genes_df <- gene_df[order(-gene_df$Total_Up_Regulated), ]
#top_10_genes <- head(top_genes_df, 10)

#print(top_10_genes)

write.table(top_genes_df, "Top_Downregulated_Genes.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
```


Visualisation heatmap UP
```{r}
gene_data_path <- "./TSV_files/General/Top_Upregulated_Genes.tsv" #Your path here
gene_df <- read.table(gene_data_path, header = TRUE, sep = "\t", check.names = FALSE)

original_gene_order <- gene_df$Gene

gene_df_long <- melt(gene_df, id.vars = "Gene", variable.name = "Cluster", value.name = "Regulated")

gene_df_long <- gene_df_long[!grepl("Total_Up_Regulated", gene_df_long$Cluster),]

gene_matrix <- dcast(gene_df_long, Gene ~ Cluster, value.var = "Regulated")
gene_matrix <- as.matrix(gene_matrix[, -1]) 
rownames(gene_matrix) <- original_gene_order 

rownames(gene_matrix) <- gene_df_long$Gene[!duplicated(gene_df_long$Gene)] 
gene_matrix <- gene_matrix[order(rownames(gene_matrix)), ]

png("enhanced_heatmap.png", width = 1600, height = 1200)

heatmap(gene_matrix, Colv = NA, Rowv = NA, scale = "none", 
        col = c("black", "green"),  
        margins = c(10, 15), 
        xlab = "Clusters", ylab = "Genes", cexRow = 0.5, cexCol = 0.5) 

dev.off()

```

Visualisation heatmap DOWN
```{r}
gene_data_path <- "./TSV_files/General/Top_Downregulated_Genes.tsv" #Your path here
gene_df <- read.table(gene_data_path, header = TRUE, sep = "\t", check.names = FALSE)

original_gene_order <- gene_df$Gene

gene_df_long <- melt(gene_df, id.vars = "Gene", variable.name = "Cluster", value.name = "Regulated")

gene_df_long <- gene_df_long[!grepl("Total_Up_Regulated", gene_df_long$Cluster),]

gene_matrix <- dcast(gene_df_long, Gene ~ Cluster, value.var = "Regulated")
gene_matrix <- as.matrix(gene_matrix[, -1]) 
rownames(gene_matrix) <- original_gene_order

rownames(gene_matrix) <- gene_df_long$Gene[!duplicated(gene_df_long$Gene)]  
gene_matrix <- gene_matrix[order(rownames(gene_matrix)), ]

png("enhanced_heatmap_down.png", width = 1600, height = 1200)  

heatmap(gene_matrix, Colv = NA, Rowv = NA, scale = "none", 
        col = c("black", "red"), 
        margins = c(10, 15),  
        xlab = "Clusters", ylab = "Genes", cexRow = 0.5, cexCol = 0.5)  

dev.off()
```






REACTOME ANALYSIS

Analyse fgsea (3.2) - Results in TSV files for each cluster
```{r}
library(fgsea)

pathways.hallmark <- gmtPathways("./TSV_files/raw/reactome.gmt") #Your path here

lapply(head(pathways.hallmark), head)

input_folder <- "./TSV_files/Cluster_filtered" #Your path here
files <- list.files(input_folder, pattern = "\\.tsv$", full.names = TRUE)
files <- sort(files)

for (i in seq_along(files)) {
  file_path <- files[i]
  data <- read.table(file_path, header = TRUE, sep = "\t")
  
  ranks <- setNames(data$stat, data$Genes)
  
  fgseaRes <- fgsea(pathways = pathways.hallmark, stats = ranks, nperm = 1000)
  
  fgseaResTidy <- as.data.frame(fgseaRes) %>% 
    mutate(across(where(is.list), ~ sapply(., toString))) %>%  
    arrange(desc(NES))

  cluster_number_formatted <- sprintf("%02d", i - 1)  

  output_path <- paste0("./TSV_files/fgsea_results/GSEA_results_Cluster_", cluster_number_formatted, ".tsv") #Your path here
  write.table(fgseaResTidy, file = output_path, sep = "\t", row.names = FALSE, quote = FALSE)
}




```


Output matrix - reactome significatif
```{r}
input_folder <- "./TSV_files/fgsea_results" #Your path here

files <- list.files(input_folder, pattern = "\\.tsv$", full.names = TRUE)
files <- sort(files)
print(files)

result_matrix_fgsea <- matrix(nrow = length(files), ncol = 2, dimnames = list(NULL, c("Cluster", "Significant_Reactomes")))

for (i in seq_along(files)) {
  file_path <- files[i]
  data <- read.table(file_path, header = TRUE, sep = "\t")

  significant_genes_count <- sum(data$padj < 0.05, na.rm = TRUE) 
  cluster_number <- as.numeric(sub(".*_Cluster_(\\d+)\\.tsv", "\\1", basename(file_path)))
  result_matrix_fgsea[i, ] <- c(cluster_number, significant_genes_count)
}

print(result_matrix_fgsea)
```

Output matrix - Reactome UP/DOWN
```{r}
nes <- 0
pval_threshold <- 0.05

input_folder <- "./TSV_files/fgsea_results" #Your path here

files <- list.files(input_folder, pattern = "\\.tsv$", full.names = TRUE)
files <- sort(files)

result_matrix_fgsea_up_down <- matrix(nrow = length(files), ncol = 3, dimnames = list(NULL, c("Cluster", "Up_reactome", "Down_reactome")))

for (i in seq_along(files)) {
  file_path <- files[i]
  data <- read.table(file_path, header = TRUE, sep = "\t")

  up_regulated_count <- sum(data$NES > nes & data$padj < pval_threshold, na.rm = TRUE)
  
  down_regulated_count <- sum(data$NES < nes & data$padj < pval_threshold, na.rm = TRUE)

  cluster_number <-as.numeric(sub(".*_Cluster_(\\d+)\\.tsv", "\\1", basename(file_path)))

  result_matrix_fgsea_up_down[i, ] <- c(cluster_number, up_regulated_count, down_regulated_count)
}

print(result_matrix_fgsea_up_down)
```


TOP matrix - reactome
```{r}
nes <- 0
pval_threshold <- 0.05

input_folder <- "./TSV_files/fgsea_results" #Your path here
files <- list.files(input_folder, pattern = "\\.tsv$", full.names = TRUE)
files <- sort(files)

genes_file_path <- "./TSV_files/fgsea_results/GSEA_results_Cluster_00.tsv" #Your path here
genes_list <- read.table(genes_file_path, header = TRUE, sep = "\t")
unique_genes <- unique(genes_list$pathway)  

gene_df <- data.frame(pathway = unique_genes)

for (file_path in files) {
    data <- read.table(file_path, header = TRUE, sep = "\t")
    
    up_regulated <- data[data$NES > nes & data$padj < pval_threshold, ]

    cluster_name <- as.numeric(sub(".*_Cluster_(\\d+)\\.tsv", "\\1", basename(file_path)))
    gene_df[[paste0("Cluster_", cluster_name)]] <- 0  

    if (length(up_regulated) > 0) {
        gene_df[[paste0("Cluster_", cluster_name)]] <- ifelse(gene_df$pathway %in% up_regulated$pathway, 1, 0)
    }
}

gene_df$Total_Up_Regulated <- rowSums(gene_df[, -1])

top_genes_df <- gene_df[order(-gene_df$Total_Up_Regulated), ]
#top_10_genes <- head(top_genes_df, 10)

#print(top_10_genes)
write.table(top_genes_df, "Top_Upregulated_Reactome.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
```


DOWN matrix - reactome
```{r}
nes <- 0
pval_threshold <- 0.05

input_folder <- "./TSV_files/fgsea_results" #Your path here 
files <- list.files(input_folder, pattern = "\\.tsv$", full.names = TRUE)
files <- sort(files)

genes_file_path <- "./TSV_files/fgsea_results/GSEA_results_Cluster_00.tsv" #Your path here
genes_list <- read.table(genes_file_path, header = TRUE, sep = "\t")
unique_genes <- unique(genes_list$pathway)  

gene_df <- data.frame(pathway = unique_genes)

for (file_path in files) {
    data <- read.table(file_path, header = TRUE, sep = "\t")
    
    up_regulated <- data[data$NES < nes & data$padj < pval_threshold, ]

    cluster_name <- as.numeric(sub(".*_Cluster_(\\d+)\\.tsv", "\\1", basename(file_path)))
    gene_df[[paste0("Cluster_", cluster_name)]] <- 0  

    if (length(up_regulated) > 0) {
        gene_df[[paste0("Cluster_", cluster_name)]] <- ifelse(gene_df$pathway %in% up_regulated$pathway, 1, 0)
    }
}

gene_df$Total_Up_Regulated <- rowSums(gene_df[, -1])

top_genes_df <- gene_df[order(-gene_df$Total_Up_Regulated), ]
#top_10_genes <- head(top_genes_df, 10)

#print(top_10_genes)

write.table(top_genes_df, "Top_Downregulated_Reactome.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
```

Visualisation heatmap UP - Reactome
```{r}
gene_data_path <- "./TSV_files/General/Top_Upregulated_Reactome.tsv" #Your path here
gene_df <- read.table(gene_data_path, header = TRUE, sep = "\t", check.names = FALSE)

original_gene_order <- gene_df$pathway

gene_df_long <- melt(gene_df, id.vars = "pathway", variable.name = "Cluster", value.name = "Regulated")

gene_df_long <- gene_df_long[!grepl("Total_Up_Regulated", gene_df_long$Cluster),]

gene_matrix <- dcast(gene_df_long, pathway ~ Cluster, value.var = "Regulated")
gene_matrix <- as.matrix(gene_matrix[, -1]) 
rownames(gene_matrix) <- original_gene_order 

rownames(gene_matrix) <- gene_df_long$pathway[!duplicated(gene_df_long$pathway)] 
gene_matrix <- gene_matrix[order(rownames(gene_matrix)), ] 

png("enhanced_heatmap_reactome.png", width = 1600, height = 1200)  

heatmap(gene_matrix, Colv = NA, Rowv = NA, scale = "none", 
        col = c("black", "green"),  
        margins = c(10, 15),  
        xlab = "Clusters", ylab = "Pathway", cexRow = 0.5, cexCol = 0.5) 

dev.off()

```

Visualisation heatmap DOWN - Reactome
```{r}
gene_data_path <- "./TSV_files/General/Top_Downregulated_Reactome.tsv" #Your path here
gene_df <- read.table(gene_data_path, header = TRUE, sep = "\t", check.names = FALSE)

original_gene_order <- gene_df$pathway

gene_df_long <- melt(gene_df, id.vars = "pathway", variable.name = "Cluster", value.name = "Regulated")

gene_df_long <- gene_df_long[!grepl("Total_Up_Regulated", gene_df_long$Cluster),]

gene_matrix <- dcast(gene_df_long, pathway ~ Cluster, value.var = "Regulated")
gene_matrix <- as.matrix(gene_matrix[, -1]) 
rownames(gene_matrix) <- original_gene_order

rownames(gene_matrix) <- gene_df_long$pathway[!duplicated(gene_df_long$pathway)]  
gene_matrix <- gene_matrix[order(rownames(gene_matrix)), ] 

png("enhanced_heatmap_reactome_down.png", width = 1600, height = 1200)  

heatmap(gene_matrix, Colv = NA, Rowv = NA, scale = "none", 
        col = c("black", "red"), 
        margins = c(10, 15),  
        xlab = "Clusters", ylab = "Pathway", cexRow = 0.5, cexCol = 0.5)  

dev.off()

```







