---
title: "Project - Group 4"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

[Link to the instructions of the project](https://owncloud.ulb.ac.be/index.php/s/an85YFknNiDQXGA?path=%2Fresources#pdfviewer)  
[Link to the information of the variables name](https://ftp.ncbi.nlm.nih.gov/dbgap/studies/phs000424/phs000424.v8.p2/pheno_variable_summaries/phs000424.v8.pht002742.v8.GTEx_Subject_Phenotypes.data_dict.xml)   
Technical variable:  
- COHORT(categorical)    
- TRISCHD(numerical)   
- DTHHRDY(categorical)    
- DTHVNT(categorical)    
Demographic/Health:  
- AGE(numerical)   
- SEX(categorical)       
- HGHT(numerical)   
- WGHT(numerical)   
- BMI(numerical)   


#Question 3


Librairies to install and import
```{r}
#Installation of the Bioconductor + Installation of the DESeq2
if (!requireNamespace('BiocManager', quietly = TRUE))
    install.packages('BiocManager')
BiocManager::install("DESeq2")
```

```{r}
#Import all libraries
library(readr)
library(ggplot2)
library(cowplot)
library(DESeq2)
library(dplyr) #install.packages("dplyr")
library(pheatmap)
library(tidyr)
library(tibble)
library(reshape2)
library(fgsea)
```

Global variables - PATH FOR COMPUTER
```{r}
general_folder_local_path <- "" #COMPLETE HERE THE PATH
```


Path definition and global values
```{r}
# Path to file (raw data)
clinical_path <- file.path(general_folder_local_path, "TSV_files", "raw", "clinical_data.tsv")
df_clinical <- read.delim(clinical_path)

morphological_path <- file.path(general_folder_local_path, "TSV_files", "raw", "morphological_counts_lunit_dino.tsv")
df_morphological <- read.delim(morphological_path)

rna_path <- file.path(general_folder_local_path, "TSV_files", "raw", "RNA_read_counts.tsv")
df_rna <- read.delim(rna_path)
```

```{r}
#Global variables
cluster_covariate_analysis <- "Mophological.cluster.G4_19"
log2fc_threshold <- 0.32 #To set the ratio of up/down regulated to 1.25
pval_threshold <- 0.05

```

```{r}
#Path to result obtained to speed up the process

input_folder <- "result_folder" 

input_folder_path <- file.path(general_folder_local_path, "TSV_files", "Result_Q3", "Result_Covariate_Cluster19_Morpho_log10_DeSeq2")
genes_file_path <- file.path(general_folder_local_path, "TSV_files", "Result_Q3", "Mophological.cluster.G4_00_result.tsv")
gene_data_up_path <- file.path(general_folder_local_path, "TSV_files", "Result_Q3", "Result_Genes", "Top_Upregulated_Genes.tsv")
gene_data_down_path <- file.path(general_folder_local_path, "TSV_files", "Result_Q3", "Result_Genes", "Top_Downregulated_Genes.tsv") 
gene_data_up_path_filtered <- file.path(general_folder_local_path, "TSV_files", "Result_Q3", "Result_Genes", "Top_Upregulated_Genes_filtered.tsv")
gene_data_down_path_filtered <- file.path(general_folder_local_path, "TSV_files", "Result_Q3", "Result_Genes", "Top_Downregulated_Genes_filtered.tsv") 
```




Cleaning process - Filtering genes
```{r}
#Cleaning PROCRESS:

#Start with 56200 genes in total
#First filter, genes that has a sum less than 10 is removed from the dataset (45675 genes left)

df_rna_clean <- df_rna %>%
  mutate(across(3:ncol(df_rna), as.numeric)) %>%
  mutate(sum_expression = rowSums(across(3:ncol(df_rna)), na.rm = TRUE)) %>%
  filter(sum_expression > 10) %>%
  select(-sum_expression)

str(df_rna_clean)



#Second filter: calcul the MAD value for each gene (Median Absolute Deviation) because we want to keep the genes with the highest deviation and variability. In this case, we are going to use the log10(MAD +1), as explained in the project question as hint. Then, based on the histogram, a threshold is found (Q3) to separate the histogram into 2 different groups, with the highest and lowest value, to remove all the genes with the lowest variations. At the end, we only have 11382 genes left.

median_expression <- apply(df_rna_clean[, 3:ncol(df_rna_clean)], 1, median, na.rm = TRUE)
mad_values <- apply(df_rna_clean[, 3:ncol(df_rna_clean)], 1, function(x) {
  median(abs(x - median(x, na.rm = TRUE)), na.rm = TRUE)
})

df_rna_clean$mad_value <- mad_values

df_rna_clean$log10_mad_plus_1 <- log10(df_rna_clean$mad_value + 1)

threshold_mad_log10 <- quantile(df_rna_clean$log10_mad_plus_1, 0.75, na.rm = TRUE)

print(dim(df_rna_clean))
print(head(df_rna_clean))

histogram_plot <- ggplot(df_rna_clean, aes(x = log10_mad_plus_1)) +
  geom_histogram(binwidth = 0.5, fill = "skyblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of log10(MAD + 1)",
       x = "log10(MAD + 1)",
       y = "Frequency") +
  geom_vline(aes(xintercept = threshold_mad_log10), color = "red", linetype = "dashed") +
  geom_text(aes(x = threshold_mad_log10, y = Inf, label = paste("Threshold =", round(threshold_mad_log10, 2))), 
            color = "red", angle = 90, vjust = 1.5, hjust = -0.5)

print(histogram_plot)

df_rna_high_variability <- df_rna_clean %>%
  filter(log10_mad_plus_1 > threshold_mad_log10) %>%
  select(-log10_mad_plus_1)

print(dim(df_rna_high_variability))
print(head(df_rna_high_variability))

# write.table(df_rna_high_variability, file = "df_rna_filtered.tsv", sep = "\t", row.names = FALSE) #Uncomment to save it

```

Transformation the df_morphological count using the log10(x+1) for each value
```{r}
df_transformed_morpho <- apply(df_morphological[, 2:length(df_morphological)], 2, function(x) log10(as.numeric(x) + 1))

df_transformed_morpho <- as.data.frame(df_transformed_morpho)

df_transformed_morpho$SMPLID <- df_morphological[, 1]

df_transformed_morpho <- df_transformed_morpho[, c("SMPLID", setdiff(names(df_transformed_morpho), "SMPLID"))]

print(df_transformed_morpho)
```


Run DeSeq analysis 
```{r}
run_deseq <- function(formula, count_matrix, col_data) {
  print(formula)
  
  dds <- DESeqDataSetFromMatrix(countData = count_matrix, colData = col_data, design = as.formula(formula))
  dds <- DESeq(dds)
  
  print(results(dds))
  return(results(dds))
}

cluster_name <- colnames(df_morphological[, 2:length(df_morphological)])
df_rna_filtered_clean <- df_rna_filtered[, 4:length(df_rna_filtered)-1]

for (cluster in cluster_name) {
  res <- run_deseq(paste0("~ ", cluster, " + ", cluster_covariate_analysis), df_rna_filtered_clean, df_morphological)
  
  res_df <- as.data.frame(res)
  res_df$Gene <- df_rna_filtered[, 2]

  res_df <- res_df[, c("Gene", setdiff(names(res_df), "Gene"))]
  #write.table(res_df, file = paste0("new_result5/", cluster, "_result.tsv"), sep = "\t", row.names = FALSE, col.names = TRUE, quote = FALSE)
}




```

Matrix UP/DOWN regulated genes + tot
```{r}
files <- list.files(input_folder_path, pattern = "\\.tsv$", full.names = TRUE)
files <- sort(files)

result_matrix <- matrix(nrow = length(files), ncol = 4, dimnames = list(NULL, c("Cluster", "Up_regulated_Genes", "Down_regulated_Genes", "Total_significant")))
count <- 0

for (i in seq_along(files)) {
  file_path <- files[i]
  data <- read.table(file_path, header = TRUE, sep = "\t")

  up_regulated_count <- sum(data$log2FoldChange > log2fc_threshold & data$padj < pval_threshold, na.rm = TRUE)
  
  down_regulated_count <- sum(data$log2FoldChange < -log2fc_threshold & data$padj < pval_threshold, na.rm = TRUE)
  
  total_significant <- up_regulated_count + down_regulated_count

  result_matrix[i, ] <- c(count, up_regulated_count, down_regulated_count, total_significant)
  
  count <- count + 1
}

print(result_matrix)

```

Creation matrix top regulated genes
```{r}
files <- list.files(input_folder, pattern = "\\.tsv$", full.names = TRUE)
files <- sort(files)

genes_list <- read.table(genes_file_path, header = TRUE, sep = "\t")
unique_genes <- unique(genes_list$Gene)  

gene_df <- data.frame(Gene = unique_genes)
count <- 0

for (file_path in files) {
    data <- read.table(file_path, header = TRUE, sep = "\t")
    
    up_regulated <- data[data$log2FoldChange > log2fc_threshold & data$padj < pval_threshold, ]
    
    gene_df[[paste0("Cluster_", count)]] <- 0 
    
    if (length(up_regulated) > 0) {
        gene_df[[paste0("Cluster_", count)]] <- ifelse(gene_df$Gene %in% up_regulated$Gene, 1, 0)
    }
    count <- count +1
}

gene_df$Total_Up_Regulated <- rowSums(gene_df[, -1])

top_genes_df <- gene_df[order(-gene_df$Total_Up_Regulated), ]
#top_10_genes <- head(top_genes_df, 10)
#print(top_10_genes)

#write.table(top_genes_df, "Top_Upregulated_Genes_test.tsv", sep = "\t", row.names = FALSE, quote = FALSE)

```

Creation matrix down regulated genes
```{r}
files <- list.files(input_folder, pattern = "\\.tsv$", full.names = TRUE)
files <- sort(files)

genes_list <- read.table(genes_file_path, header = TRUE, sep = "\t")
unique_genes <- unique(genes_list$Gene)  

gene_df <- data.frame(Gene = unique_genes)
count <- 0

for (file_path in files) {
    data <- read.table(file_path, header = TRUE, sep = "\t")
    
    down_regulated <- data[data$log2FoldChange < -log2fc_threshold & data$padj < pval_threshold, ]

    gene_df[[paste0("Cluster_", count)]] <- 0  

    if (length(down_regulated) > 0) {
        gene_df[[paste0("Cluster_", count)]] <- ifelse(gene_df$Gene %in% down_regulated$Gene, 1, 0)
    }
    
    count <- count + 1
}

gene_df$Total_Up_Regulated <- rowSums(gene_df[, -1])

top_genes_df <- gene_df[order(-gene_df$Total_Up_Regulated), ]
#top_10_genes <- head(top_genes_df, 10)

#print(top_10_genes)

#write.table(top_genes_df, "Top_Downregulated_Genes_test.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
```

Visualisation heatmap UP
```{r}
gene_df <- read.table(gene_data_up_path, header = TRUE, sep = "\t", check.names = FALSE)

original_gene_order <- gene_df$Gene

gene_df_long <- melt(gene_df, id.vars = "Gene", variable.name = "Cluster", value.name = "Regulated")

gene_df_long <- gene_df_long[!grepl("Total_Up_Regulated", gene_df_long$Cluster),]

gene_matrix <- dcast(gene_df_long, Gene ~ Cluster, value.var = "Regulated")
gene_matrix <- as.matrix(gene_matrix[, -1]) 
rownames(gene_matrix) <- original_gene_order 

rownames(gene_matrix) <- gene_df_long$Gene[!duplicated(gene_df_long$Gene)] 
gene_matrix <- gene_matrix[order(rownames(gene_matrix)), ]

png("heatmap_up.png", width = 1600, height = 1200)

heatmap(gene_matrix, Colv = NA, Rowv = NA, scale = "none", 
        col = c("black", "green"),  
        margins = c(10, 15), 
        xlab = "Clusters", ylab = "Genes", cexRow = 0.5, cexCol = 0.5) 

dev.off()

```

Visualisation heatmap DOWN
```{r}
gene_df <- read.table(gene_data_down_path, header = TRUE, sep = "\t", check.names = FALSE)

original_gene_order <- gene_df$Gene

gene_df_long <- melt(gene_df, id.vars = "Gene", variable.name = "Cluster", value.name = "Regulated")

gene_df_long <- gene_df_long[!grepl("Total_Up_Regulated", gene_df_long$Cluster),]

gene_matrix <- dcast(gene_df_long, Gene ~ Cluster, value.var = "Regulated")
gene_matrix <- as.matrix(gene_matrix[, -1]) 
rownames(gene_matrix) <- original_gene_order

rownames(gene_matrix) <- gene_df_long$Gene[!duplicated(gene_df_long$Gene)]  
gene_matrix <- gene_matrix[order(rownames(gene_matrix)), ]

png("heatmap_downtest.png", width = 1600, height = 1200)  

heatmap(gene_matrix, Colv = NA, Rowv = NA, scale = "none", 
        col = c("black", "red"), 
        margins = c(10, 15),  
        xlab = "Clusters", ylab = "Genes", cexRow = 0.5, cexCol = 0.5)  

dev.off()
```







REACTOME ANALYSIS
Path definition
```{r}
reactom_gmt_path = file.path(general_folder_local_path, "TSV_files", "raw", "reactome.gmt")
pathways.hallmark <- gmtPathways(reactom_gmt_path) 


input_folder_fgsea <- "fgsea_results" 

input_folder_fgsea_path <- file.path(general_folder_local_path, "TSV_files", "Result_Q3", "fgsea_results")

output_fgsea_results_path <- file.path(general_folder_local_path, "TSV_files", input_folder_fgsea, "GSEA_results_Cluster_")

#File already created, to speed up the time process running
reactome_file_path <- file.path(general_folder_local_path, "TSV_files", input_folder_fgsea, "GSEA_results_Cluster_00.tsv")
reactome_data_up_path <- file.path(general_folder_local_path, "TSV_files", "Result_Q3", "Result_Reactome", "Top_Upregulated_Reactome.tsv" )
reactome_data_down_path <- file.path(general_folder_local_path, "TSV_files", "Result_Q3", "Result_Reactome", "Top_Downregulated_Reactome.tsv" )
```

Global variables
```{r}
nes <- 2
pval_threshold <- 0.05
```


Analyse fgsea (3.2) - Results in TSV files for each cluster
```{r}
lapply(head(pathways.hallmark), head)

files <- list.files(input_folder, pattern = "\\.tsv$", full.names = TRUE)
print(input_folder)
files <- sort(files)

for (i in seq_along(files)) {
  file_path <- files[i]
  data <- read.table(file_path, header = TRUE, sep = "\t")
  
  ranks <- setNames(data$stat, data$Gene)
  
  fgseaRes <- fgsea(pathways = pathways.hallmark, stats = ranks, nperm = 1000)
  
  fgseaResTidy <- as.data.frame(fgseaRes) %>% 
    mutate(across(where(is.list), ~ sapply(., toString))) %>%  
    arrange(desc(NES))

  cluster_number_formatted <- sprintf("%02d", i - 1)  

  output_path <- paste0(output_fgsea_results_path, cluster_number_formatted, ".tsv") #Your path here
  write.table(fgseaResTidy, file = output_path, sep = "\t", row.names = FALSE, quote = FALSE)
}

```

Output matrix - Reactome UP/DOWN
```{r}
files <- list.files(input_folder_fgsea_path, pattern = "\\.tsv$", full.names = TRUE)
files <- sort(files)

result_matrix_fgsea_up_down <- matrix(nrow = length(files), ncol = 4, dimnames = list(NULL, c("Cluster", "Up_reactome", "Down_reactome", "Total_significant")))

count <- 0

for (i in seq_along(files)) {
  file_path <- files[i]
  data <- read.table(file_path, header = TRUE, sep = "\t")

  up_regulated_count <- sum(data$NES > nes & data$padj < pval_threshold, na.rm = TRUE)
  
  down_regulated_count <- sum(data$NES < -nes & data$padj < pval_threshold, na.rm = TRUE)
  
  total_significant <- up_regulated_count + down_regulated_count

  result_matrix_fgsea_up_down[i, ] <- c(count, up_regulated_count, down_regulated_count, total_significant)
  count <- count +1
}

print(result_matrix_fgsea_up_down)
```

TOP matrix - reactome
```{r}
files <- list.files(input_folder_fgsea, pattern = "\\.tsv$", full.names = TRUE)
files <- sort(files)

genes_list <- read.table(reactome_file_path, header = TRUE, sep = "\t")
unique_genes <- unique(genes_list$pathway)  

gene_df <- data.frame(pathway = unique_genes)
count <- 0

for (file_path in files) {
    data <- read.table(file_path, header = TRUE, sep = "\t")
    
    up_regulated <- data[data$NES > nes & data$padj < pval_threshold, ]

    gene_df[[paste0("Cluster_", count)]] <- 0  

    if (length(up_regulated) > 0) {
        gene_df[[paste0("Cluster_", count)]] <- ifelse(gene_df$pathway %in% up_regulated$pathway, 1, 0)
    }
    count <- count +1
}

gene_df$Total_Up_Regulated <- rowSums(gene_df[, -1])

top_genes_df <- gene_df[order(-gene_df$Total_Up_Regulated), ]
#top_10_genes <- head(top_genes_df, 10)

#print(top_10_genes)
#write.table(top_genes_df, "Top_Upregulated_Reactome_test.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
```

DOWN matrix - reactome
```{r}
files <- list.files(input_folder_fgsea, pattern = "\\.tsv$", full.names = TRUE)
files <- sort(files)

genes_list <- read.table(reactome_file_path, header = TRUE, sep = "\t")
unique_genes <- unique(genes_list$pathway)  

gene_df <- data.frame(pathway = unique_genes)
count <- 0

for (file_path in files) {
    data <- read.table(file_path, header = TRUE, sep = "\t")
    
    up_regulated <- data[data$NES < -nes & data$padj < pval_threshold, ]

    gene_df[[paste0("Cluster_", count)]] <- 0  

    if (length(up_regulated) > 0) {
        gene_df[[paste0("Cluster_", count)]] <- ifelse(gene_df$pathway %in% up_regulated$pathway, 1, 0)
    }
    count <- count +1
}

gene_df$Total_Up_Regulated <- rowSums(gene_df[, -1])

top_genes_df <- gene_df[order(-gene_df$Total_Up_Regulated), ]
#top_10_genes <- head(top_genes_df, 10)

#print(top_10_genes)

#write.table(top_genes_df, "Top_Downregulated_Reactome_test.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
```

Visualisation heatmap UP - Reactome
```{r}
gene_df <- read.table(reactome_data_up_path, header = TRUE, sep = "\t", check.names = FALSE)

original_gene_order <- gene_df$pathway

gene_df_long <- melt(gene_df, id.vars = "pathway", variable.name = "Cluster", value.name = "Regulated")

gene_df_long <- gene_df_long[!grepl("Total_Up_Regulated", gene_df_long$Cluster),]

gene_matrix <- dcast(gene_df_long, pathway ~ Cluster, value.var = "Regulated")
gene_matrix <- as.matrix(gene_matrix[, -1]) 
rownames(gene_matrix) <- original_gene_order 

rownames(gene_matrix) <- gene_df_long$pathway[!duplicated(gene_df_long$pathway)] 
gene_matrix <- gene_matrix[order(rownames(gene_matrix)), ] 

png("heatmap_up_reactome.png", width = 1600, height = 1200)  

heatmap(gene_matrix, Colv = NA, Rowv = NA, scale = "none", 
        col = c("black", "green"),  
        margins = c(10, 15),  
        xlab = "Clusters", ylab = "Pathway", cexRow = 0.5, cexCol = 0.5) 

dev.off()

```

Visualisation heatmap DOWN - Reactome
```{r}
gene_df <- read.table(reactome_data_down_path, header = TRUE, sep = "\t", check.names = FALSE)

original_gene_order <- gene_df$pathway

gene_df_long <- melt(gene_df, id.vars = "pathway", variable.name = "Cluster", value.name = "Regulated")

gene_df_long <- gene_df_long[!grepl("Total_Up_Regulated", gene_df_long$Cluster),]

gene_matrix <- dcast(gene_df_long, pathway ~ Cluster, value.var = "Regulated")
gene_matrix <- as.matrix(gene_matrix[, -1]) 
rownames(gene_matrix) <- original_gene_order

rownames(gene_matrix) <- gene_df_long$pathway[!duplicated(gene_df_long$pathway)]  
gene_matrix <- gene_matrix[order(rownames(gene_matrix)), ] 

png("heatmap_down_reactome.png", width = 1600, height = 1200)  

heatmap(gene_matrix, Colv = NA, Rowv = NA, scale = "none", 
        col = c("black", "red"), 
        margins = c(10, 15),  
        xlab = "Clusters", ylab = "Pathway", cexRow = 0.5, cexCol = 0.5)  

dev.off()

```



